---
title: "Gas Exploration Induced Earthquakes"
subtitle: "Groningen Gas Field, The Netherlands"
author: "Maaike Miedema"
date: "June 16, 2017"
output:
  ioslides_presentation: default
  slidy_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r libraries}
### LOAD LIBRARIES ###
library(dplyr)
library(stringr)
library(plotly)
library(lubridate)
library(RColorBrewer)

```
```{r load_data}
### LOAD DATA ###
# source: http://knmi.nl/kennis-en-datacentrum/dataset/aardbevingscatalogus, 
# file all_induced.csv downloaded June 14, 2017
allData<-read.csv("all_induced.csv")
```

```{r clean_data }
### CLEAN DATA ###
# remove negative magnitudes
# convert date and time 

fill_out <- function(string){
        # for times 00:00:00 to 09:59:59 time presentation needs one 
        # or more zero's in front to process to a time
        n <- 6-str_length(string)
        while(n>0){
                string<-paste("0", string, sep="")
                n<-n-1
        }
        string
}

earthquake <- allData %>%
        select(-EVALMODE)%>%
        filter(MAG>0, LON>5.4)
earthquake$TIME <-  earthquake$TIME %>% 
        as.integer()%>%
        as.character()
earthquake$TIME <- lapply(earthquake$TIME, fill_out)
earthquake <-mutate(earthquake, DATE_TIME= paste(YYMMDD, TIME, sep=""))
earthquake$DATE_TIME <-ymd_hms(earthquake$DATE_TIME)
```
```{r factors}
### FACTORIZE VARIABLES FOR HEATMAP PLOT
earthquake <- mutate(earthquake, MAGFACTOR=MAG)
earthquake$MAGFACTOR <- cut(earthquake$MAGFACTOR, breaks = 0.5*(0:8))
earthquake <- mutate(earthquake, YEAR =year(DATE_TIME))



# matrix for heatmap

grouped<-group_by(earthquake, YEAR, MAG)
number<-group_size(grouped)
df<-data.frame(summarize(grouped), NUMBER=number)
plotheat<-matrix(nrow=length(1986:2017), ncol=length(0:36), 0)
rownames(plotheat)<-as.character(1986:2017)
colnames(plotheat)<-as.character(0.1*(0:36))
df<-mutate(df,YEAR=as.character(YEAR), MAG=as.character(MAG))

fill_heat<-function(df, heat){
        df<-arrange(df, YEAR, MAG)
        k<-1
        for (k in 1:dim(df)[1]){
        heat[df$YEAR[k],df$MAG[k]]<-df$NUMBER[k]
        }
        heat
}

plotheat<-fill_heat(df,plotheat)

```


## Introduction

The Groningen gas field in the North Eastern part of the Netherlands is exploited since 1963.   
This gas exploration leads to subsidience and earthquakes, causing damage to buildings. 

The quakes happen just below surface (+/- 3 km), therefore  
damage to buildings is bigger than expected considering their magnitude alone.

This presentation shows when and with what magnitude earthquakes have happened. 

## Dutch Earthquake Data

- Collected and distributed by the Royal Netherlands Meteorologic Institute.  
- Describes: location, time, depth and magnitude. 
- Three decades: from 1986-12-26 to 2017-06-14.

We only consider data of earthquakes induced by gas exploration near Groningen


## Earthquakes the last 30 years

```{r scatterplot}
plot_ly(earthquake,x =~DATE_TIME, y=~MAG, type="scatter") %>%
        layout(yaxis=list(title="Magnitude (Richter)"), xaxis=list(title="Time"))
```
As you can see: as time goes by more and more quakes happen.


## Earthquakes per year and magnitude

```{r heatmap}
plot_ly(x=colnames(plotheat),y=rownames(plotheat),z=~plotheat, type = "heatmap", 
        colors="Spectral",colorbar=list(title="Number",len=0.4))%>%
        layout(yaxis=list(title="Year"),xaxis=list(title="Magnitude (Richter)"))
```

  
Remark: incomplete 2017 data, we are halfway the year.

## Information

- <b> Data:</b> all_induced.csv  
from <a href="http://knmi.nl/kennis-en-datacentrum/dataset/aardbevingscatalogus">KNMI</a>,
download June 14, 2017.
- <b>Damage</b>, for an impression see: <a href=" http://www.dwarshuis.com/aardbevingen-groningen/zo-veel-mogelijk-fotos-van-de-aardbevings-schade/2.php">Photos</a>
- <b> Code of slides</b> can be found at:  
<a href=https://github.com/MaaikeMiedema/EarthquakesPresentation> github.com/MaaikeMiedema/EarthquakesPresentation </a>

This presentation is made as an assignment for the Data Products Development class on Coursera.
